Dark Control:\\ A Unified Account of Default Mode Function\\ by Control Theory and Reinforcement Learning

### 2017-05-06 paper/paper.tex
Major pass

In artificial intelligence and machine learning, a popular computational model for
multi-step decision processes in such an environment are
\textit{Markov decision processes} (MDPs) \citep{sutton1998reinforcement}.
% For the remainder of this paper, for the sake of simplicity
% we use the abbreviation ``MDP'' to refer to the special case
% of \textit{Partially Observable Markov Decision Processes}.
An MDP operationalizes a sequential decision process
in which it is assumed that environment dynamics are determined by a Markov process,
but the agent cannot directly observe the underlying state.
Instead, the agent tries to optimize a \textit{subjective} reward
signal (i.e., is likely to be different for another agent in the same state).
This is performed alongside each transition in the state space of the environment
 and maintains a probability distribution over the set of possible
states, based on a set of observations and observation probabilities.
This is a minimal set of assumptions that can be made about an environment,
and is characteristic of so-called \textit{model-free RL}. Model-free RL has had great success in real-world problems including robotics
\citep{abbeel2004,andrewng}, computer game play, \citep{mnih2015,silver2016mastering}, etc., just to name a few.
We propose to use such a model as formal basis to model DMN function.

Such a learning scheme can be plausibly realized in the human brain \citep{doherty2015structure}. Indeed,
a core property of human intelligence has been proposed to underly improvement
of expected utility outcomes as a strategy for action choice in uncertain
environments \citep{gershman2015computational}, a situation perfectly captured by the formalism of MDPs.
It has also long been proposed \citep{dayan2008decision} that there
is a rather direct mapping of model-free RL learning algorithms
onto aspects of the brain.
The neurotransmitter dopamine could serve
as a ``teaching signal'' to better estimate value associations
and action policies by controlling
synaptic plasticity in the reward system, including the NAc.
In contrast, \textit{model-based RL} would start off with some mechanistic assumptions about the dynamics of the world.
These can be assumptions about the physical laws governing the agent's environment or constraints on the state space and transitions between states.
% For instance, if a rat in a maze knows
% that standing still will produce no change in the environment, and in particular will not eventually lead to finding the food.
%

@@

\begin{definition*}
Mathematically, an MDP is simply a quintuplet $(\mathcal S, \mathcal A, r, p)$ where
\begin{itemize}
\item $\mathcal S$ is the set of states, such as $\mathcal S = \{\text{happy}, \text{sad}\}$.
\item $\mathcal A$ is the set of actions, such as $\mathcal A = \{\text{read}, \text{run},
  \text{laugh}, \text{sympathize}, \text{empathize}\}.$
\item $r : \mathcal S \times \mathcal A \rightarrow \mathbb R$ is the \textit{reward function},
   so that $r(s, a)$ is the instant reward for taking action $a$ in state $s$.
% \item $\mu: \mathcal S \rightarrow [0, 1]$ is the prior probability on the states so that
%   $\mu(s)$ is the probability that the environment starts off in state $s$.
\item $p : \mathcal S \times \mathcal A \times \mathcal S \rightarrow [0, 1],\; (s,a,s') \mapsto p(s'|s,a)$,
  the probability of moving to state $s'$ if action $a$ is taken from state $s$. In addition, one requires that such
  transitions be Markovian. Consequently, the future states are independent of past states and only depend on the present state and action taken.

@@

\subsubsection{Accumulated rewards and policies}
The behavior of the agent is governed by a \textit{policy}, which maps states of the world
to probability distributions over actions. Starting at time $t=0$,
followung a policy $\pi$ generates a trajectory of action choices as follows:
\begin{eqnarray*}
  \begin{split}
    \textbf{choose action: }a_0 &\sim \pi(a|s_0)\\
    \textbf{ observe transition: }s_1 &\sim p(s|s_0,a_0)\textbf{ and collect reward }R_0 = r(s_0, a_0)\\
    \textbf{choose action: }a_1 &\sim \pi(a|s_1)\\
    \textbf{ observe transition: }s_2 &\sim p(s|s_1,a_1), \textbf{ and collect reward }R_1 = r(s_1, a_1)\\
    \vdots\\
    \textbf{choose action: }a_{t} &\sim \pi(a|s_{t})\\
    \textbf{ observe transition: }s_{t+1} &\sim p(s|s_{t},a_{t}), \textbf{ and collect reward }R_{t} = r(s_{t}, a_{t})\\
    \vdots
  \end{split}
\end{eqnarray*}
We assume time-invariance
in that we expect the dynamics of the process
to be equivalent over sufficiently long time windows of equal length (i.e., stationarity).

@@

The goal of the RL agent is then to update this action policy in order
to maximize $G^\pi$ on average (cf. below). In \eqref{eq:cumr},
the definition of cumulative reward $G^\pi$,
the constant $\gamma$ $(0 \le \gamma < 1$) is the \textit{reward discount factor},
viewed to be characteristic for a certain agent.
On the one hand,
setting $\gamma = 0$ yields perfectly hedonistic behavior.
An agent with such a shortsighted horizon is exclusively
concerned with immediate rewards.
This is however not compatible with coordinated planning if long-term goal that is
potentially subserved by neural activity in the DMN.
% \footnote{In which case we can take $T \rightarrow +\infty$, provided the reward function $r$ is bounded.}.
% The reward discount factor
On the other hand,
setting $0 < \gamma < 1$ allows a learning process to arise.
A positive $\gamma$ can be seen as calibrating risk-seeking trait of the intelligent agent,
that is, the behavioral predispositions related to trading longer delays for higher rewards.
Such an agent puts relatively more emphasis on rewards expected in
a longer-term future. 
Precisely, rewards that are not expected to come within $\tau := 1/(1 - \gamma)$ time steps from the present point are disregarded.
This reduces the variance of expected rewards accumulated across
considered action cascades by limiting the depth of the search tree.
Given that there is more uncertainty in the farsighted future,
it is important to appreciate that a stochastic policy estimation
is more advantageous in many RL settings.
% For instance, a deterministic strategy in playing rock-paper-scissors can
% be easily exploited and a uniformly random choice is optimal.

@@

\begin{equation}
  \pi^{\text{greedy}}(s) = \argmax_{a \in \mathcal A}\tilde{Q}(s, a|\theta),
  \label{eq:policy}
\end{equation}
This defines a mapping from states directly to actions.
% $a \mapsto \tilde{Q}(a,s|\theta)$ of the $Q$-value function ``frozen'' at state $s$.
For instance, a simple linear model with a kernel $\phi$ would be of the
form $\tilde{Q}(s, a|\theta) = \phi(s,a)^T\theta$, where
$\phi(s,a)$ would represent a high-level representation of the state-action pairs
$(s,a)$, as was done proposed \citep{songNIPS2016}, or a eural network as recently proposed by in the
breakthrough works
\citep{mnih2015,silver2016mastering} for playing complex games (atari, Go, etc.) at superhuman levels.
%  and Parr et al. 2003 (JMLR)].
In the DMN, the dmPFC would implement such a hard-max lookup
over the action space.
The model
  parameters $\theta$ would correspond to synaptic weights and connection strengths between
  brain regions. It is a time-varying neuronal program which dictates how to move from world states $s$ to actions $a$ via the hard-max policy \eqref{eq:policy}.
  The oracle Q-values $\tilde{Q}(s, a|\theta)$ tell the DMN the (expected) usefulness of taking an action a in state $s$.
  The DMN, and in particular its dmPFC node, could then contribute to the choice, at a given state $s$, of an action $a$ which maximizes this tabulated values. This mapping from states to actions is conventionally called \textit{policy matrix}.
  Learning consists in starting from a given table and
  updating it during action choices,
  which take the agent to different table entries.

@@

% \paragraph{Parameter update.}
\begin{equation}
  \delta= -\alpha_k \nabla_{\theta_{k}}\mathcal L(\theta_{k})
  = -\alpha_k\mathbb E_{(s, a, r, s') \sim \mathcal D}[\underbrace{(\tilde{Q}(s, a|\theta_{k}) - y_k)}_{\text{prediction error}}
    \underbrace{\nabla_{\theta_{k}}\tilde{Q}(s, a|\theta_{k})}_{\text{aversion}}],
  \label{eq:oracle}
\end{equation}
where the positive constants $\alpha_1, \alpha_2,\ldots$ are learning rates.
Thus, the next action is taken to drive reward prediction errors
(or \textit{regret}) to percolate from lower to higher processing layers to modulate the
choice of future actions. It is a standard result that under special conditions on the learning rates $\alpha_k$
--namely that the sum $\sum_{k=0}^\infty\alpha_k$
diverges while $\sum_{k=0}^\infty\alpha_k^2$ converges, over time (i.e., in the limit $k \rightarrow \infty$)
the thus generated approximating sequence of $Q$-value functions
% \footnote{The proof is standard and relies on the Banach Fixed-Point Theorem applied to the Bellman transform.}
$$\tilde{Q}(.,.|\theta_0) \rightarrow \tilde{Q}(.,.|\theta_1) \rightarrow \tilde{Q}(.,.|\theta_2) \rightarrow \ldots$$
are attracted and absorbed by the optimal $Q$-value function $Q^*$ defined implicitly by the Bellman equation \eqref{eq:bellman}.

@@

  From the perspective of our model-free RL agent,
  \textit{inference} in the human brain reduces to
  generalization of
  policy and value adaptations from sampled experiences to
  successful action choices and reward predictions in future states.
  As such,
  plasticity in the DMN arises naturally:
  If an agent behaving optimally in a given environment moves
  to new, yet unexperienced environments, reward prediction errors will
  massively increase \textbf{Please check/simplify sentence}.
  This will lead to adaptation of the policy until the system converges to a
  new steady-state of optimal action decisions in volatile environments.

>>>>

In artificial intelligence and machine learning, a popular computational model for
multi-step decision processes in such an environment are
\textit{Markov decision processes} (MDPs) \citep{sutton1998reinforcement}.
% For the remainder of this paper, for the sake of simplicity
% we use the abbreviation ``MDP'' to refer to the special case
% of \textit{Partially Observable Markov Decision Processes}.
An MDP operationalizes a sequential decision process
in which it is assumed that environment dynamics are determined by a Markov process,
but the agent cannot directly observe the underlying state.
Instead, the agent tries to optimize a \textit{subjective} reward
signal (i.e., is likely to be different for another agent in the same state).
This is performed alongside each transition in the state space of the environment
 and maintains a probability distribution over the set of possible
states, based on a set of observations and observation probabilities.
This is a minimal set of assumptions that can be made about an environment,
and is characteristic of so-called \textit{model-free RL}. Model-free RL has had great success in real-world problems including robotics
\citep{ng2004,abbeel2004}, super-human play in complex games like Atari\citep{mnih2015} and Go\citep{silver2016mastering}, etc., just to name a few.
We propose to use such a model as formal basis to model DMN function.

Model-free RL can be plausibly realized in the human brain \citep{doherty2015structure}. Indeed,
a core property of human intelligence has been proposed to underly improvement
of expected utility outcomes as a strategy for action choice in uncertain
environments \citep{gershman2015computational}, a situation perfectly captured by the formalism of MDPs.
It has also long been proposed \citep{dayan2008decision} that there
is a rather direct mapping of model-free RL learning algorithms
onto aspects of the brain.
The neurotransmitter dopamine could serve
as a ``teaching signal'' to better estimate value associations
and action policies by controlling
synaptic plasticity in the reward system, including the NAc.
In contrast, \textit{model-based RL} would start off with some mechanistic assumptions about the dynamics of the world.
These can be assumptions about the physical laws governing the agent's environment or constraints on the state space and transitions between states.
% For instance, if a rat in a maze knows
% that standing still will produce no change in the environment, and in particular will not eventually lead to finding the food.
%

@@

\begin{definition*}
Mathematically, an MDP is simply a quintuplet $(\mathcal S, \mathcal A, r, p)$ where
\begin{itemize}
\item $\mathcal S$ is the set of states, such as $\mathcal S = \{\text{happy}, \text{sad}\}$.
\item $\mathcal A$ is the set of actions, such as $\mathcal A = \{\text{read}, \text{run},
  \text{laugh}, \text{sympathize}, \text{empathize}\}.$
\item $r : \mathcal S \times \mathcal A \times \mathcal S \rightarrow \mathbb R$ is the \textit{reward function},
   so that $r(s, a, s')$ is the instant reward for taking action $a$ in state $s$ followed by a state-transition $s \rightarrow s'$.
% \item $\mu: \mathcal S \rightarrow [0, 1]$ is the prior probability on the states so that
%   $\mu(s)$ is the probability that the environment starts off in state $s$.
\item $p : \mathcal S \times \mathcal A \times \mathcal S \rightarrow [0, 1],\; (s,a,s') \mapsto p(s'|s,a)$,
  the probability of moving to state $s'$ if action $a$ is taken from state $s$. In addition, one requires that such
  transitions be Markovian. Consequently, the future states are independent of past states and only depend on the present state and action taken.

@@

\subsubsection{Accumulated rewards and policies}
The behavior of the agent is governed by a \textit{policy}, which maps states of the world
to probability distributions over actions. Starting at time $t=0$,
followung a policy $\pi$ generates a trajectory of action choices as follows:
\begin{eqnarray*}
  \begin{split}
    \textbf{choose action: }a_0 &\sim \pi(a|s_0)\\
    \textbf{ observe transition: }s_1 &\sim p(s|s_0,a_0)\textbf{ and collect reward }R_0 = r(s_0, a_0, s_1)\\
    \textbf{choose action: }a_1 &\sim \pi(a|s_1)\\
    \textbf{ observe transition: }s_2 &\sim p(s|s_1,a_1), \textbf{ and collect reward }R_1 = r(s_1, a_1,s_2)\\
    \vdots\\
    \textbf{choose action: }a_{t} &\sim \pi(a|s_{t})\\
    \textbf{ observe transition: }s_{t+1} &\sim p(s|s_{t},a_{t}), \textbf{ and collect reward }R_{t} = r(s_{t}, a_{t},s_{t+1})\\
    \vdots
  \end{split}
\end{eqnarray*}
We assume time-invariance
in that we expect the dynamics of the process
to be equivalent over sufficiently long time windows of equal length (i.e., stationarity).

@@

The goal of the RL agent is then to update this action policy in order
to maximize $G^\pi$ on average (cf. below). In \eqref{eq:cumr},
the definition of cumulative reward $G^\pi$,
the constant $\gamma$ $(0 \le \gamma < 1$) is the \textit{reward discount factor},
viewed to be characteristic for a certain agent.
On the one hand,
setting $\gamma = 0$ yields perfectly hedonistic behavior.
An agent with such a shortsighted horizon is exclusively
concerned with immediate rewards.
This is however not compatible with coordinated planning if long-term goal that is
potentially subserved by neural activity in the DMN.
% \footnote{In which case we can take $T \rightarrow +\infty$, provided the reward function $r$ is bounded.}.
% The reward discount factor
On the other hand,
setting $0 < \gamma < 1$ allows a learning process to arise.
A positive $\gamma$ can be seen as calibrating risk-seeking trait of the intelligent agent,
that is, the behavioral predispositions related to trading longer delays for higher rewards.
Such an agent puts relatively more emphasis on rewards expected in
a longer-term future. 
Precisely, rewards that are not expected to come within\footnote{\textbf{N.B.:} $\frac{1}{1-\gamma} = 1 + \gamma + \gamma^2 + \gamma^3 + \ldots$}
$\tau := 1/(1 - \gamma)$
time steps from the present point are disregarded. This reduces the variance of expected rewards accumulated across
considered action cascades by limiting the depth of the search tree.
Given that there is more uncertainty in the farsighted future,
it is important to appreciate that a stochastic policy estimation
is more advantageous in many RL settings.
% For instance, a deterministic strategy in playing rock-paper-scissors can
% be easily exploited and a uniformly random choice is optimal.

@@

\begin{equation}
  \pi^{\text{greedy}}(s) = \argmax_{a \in \mathcal A}\tilde{Q}(s, a|\theta),
  \label{eq:policy}
\end{equation}
This defines a mapping from states directly to actions.
% $a \mapsto \tilde{Q}(a,s|\theta)$ of the $Q$-value function ``frozen'' at state $s$.
For instance, a simple linear model with a kernel $\phi$ would be of the
form $\tilde{Q}(s, a|\theta) = \phi(s,a)^T\theta$, where
$\phi(s,a)$ would represent a high-level representation of the state-action pairs
$(s,a)$, as was done proposed \citep{songNIPS2016}, or a eural network as recently proposed by in the
breakthrough works
\citep{mnih2015,silver2016mastering} for playing complex games (atari, Go, etc.) at superhuman levels.
%  and Parr et al. 2003 (JMLR)].
In the DMN, the dmPFC would implement such a hard-max lookup
over the action space.
The model
  parameters $\theta$ would correspond to synaptic weights and connection strengths between
  brain regions. It is a time-varying neuronal program which dictates how to move from world states $s$ to actions $a$ via the hard-max policy \eqref{eq:policy}.
  The approximating $Q$-value function $\tilde{Q}(s, a|\theta)$ would tell the DMN the (expected) usefulness of taking an action $a$ in state $s$.
  The DMN, and in particular its dmPFC node, could then contribute to the choice, at a given state $s$, of an action $a$ which maximizes these approximate
  $Q$-values. Such a mapping from states to actions is conventionally called \textit{policy matrix}\citep{mnih2015,silver2016mastering}.
  Learning consists in starting from a given table and
  updating it during action choices,
  which take the agent to different table entries.

@@

% \paragraph{Parameter update.}
\begin{equation}
  \delta= -\alpha_k \nabla_{\theta_{k}}\mathcal L(\theta_{k})
  = -\alpha_k\mathbb E_{(s, a, r, s') \sim \mathcal D}[\underbrace{(\tilde{Q}(s, a|\theta_{k}) - y_k)}_{\text{prediction error}}
    \underbrace{\nabla_{\theta_{k}}\tilde{Q}(s, a|\theta_{k})}_{\text{aversion}}],
  \label{eq:oracle}
\end{equation}
where the positive constants $\alpha_1, \alpha_2,\ldots$ are learning rates.
Thus, the next action is taken to drive reward prediction errors
(or \textit{regret}) to percolate from lower to higher processing layers to modulate the
choice of future actions. It is a standard result that under special conditions on the learning rates $\alpha_k$
(namely that the sum $\sum_{k=0}^\infty\alpha_k$ diverges while $\sum_{k=0}^\infty\alpha_k^2$ converges) over time (i.e., in the limit $k \rightarrow \infty$)
the thus generated approximating sequence of $Q$-value functions
% \footnote{The proof is standard and relies on the Banach Fixed-Point Theorem applied to the Bellman transform.}
$$\tilde{Q}(.,.|\theta_0) \rightarrow \tilde{Q}(.,.|\theta_1) \rightarrow \tilde{Q}(.,.|\theta_2) \rightarrow \ldots$$
are attracted and absorbed by the optimal $Q$-value function $Q^*$ defined implicitly by the Bellman equation \eqref{eq:bellman}.

@@

  From the perspective of our model-free RL agent,
  \textit{inference} in the human brain reduces to
  generalization of
  policy and value adaptations from sampled experiences to
  successful action choices and reward predictions in future states.
  As such,
  plasticity in the DMN arises naturally:
  if an agent behaving optimally in a given environment moves
  to new, yet unexperienced environments, reward prediction errors will
  massively increase \textbf{Please check/simplify sentence}.
  This will lead to adaptation of the policy until the system converges to a
  new steady-state of optimal action decisions in volatile environments.
